---
title: "STA303 Assignment 2Question 1"
author: "Talia Fabregas"
date: "2025-03-27"
output: pdf_document
---

```{r pmisc, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
# if pmisc is more than a week old, fetch the latest 
if(!requireNamespace("Pmisc", quietly=TRUE)) {
	install.packages("Pmisc", repos='http://r-forge.r-project.org')
}
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
library(Hmisc)
library(Pmisc)
# install.packages("mcgv")
library(mgcv)
library(kableExtra)
library(janitor)
```

# Provided code

```{r echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# retrieving the data
theUrl = "http://pbrown.ca/teaching/appliedstats/data/motorcycle.rds"
theFile = basename(theUrl)
if (!file.exists(theFile)) download.file(theUrl, theFile)
x = readRDS(theFile)

# preparing the data
x$dateInt = as.integer(x$date)
x$logMonthDays = log(Hmisc::monthDays(x$date))
x$month = factor(format(x$date, "%b"), levels = format(ISOdate(2000,
+   1:12, 1), "%b"))
res = glm(killed ~ offset(logMonthDays) + dateInt + month, data = x, 
          family = poisson(link = "log")
          )
newdata = data.frame(date = seq(as.Date("1975/1/1"), as.Date("2030/1/1"), by = "month"))
newdata$dateInt = as.integer(newdata$date)
newdata$logMonthDays = log(30)
newdata$month = "Mar"
pred1 = predict(res, newdata)
newdata$month = format(newdata$date, "%b")
pred2 = predict(res, newdata)
plot(x$date, x$killed, cex = 0.2, log = "y", xlab = "", ylab = "")
matlines(newdata$date, exp(cbind(pred1, pred2)), lty = 1)
```

# Part 1

Write down, in equations not R code, a generalized additive model suitable for this problem. Explain
each of the parts of the model and give a rationale for them (i.e. “The response variable is Gamma
distributed because the number of deaths must be positive”). (4 points)

A generalized additive model (GAM) suited for this problem is the Negative Binomial GAM. We chose a Negative Binomial because our response variable, $Y_i$ (number of motorcycle deaths in month $i$) is a non-negative count variable, and unlike a Poission, it can account for over-dispersion (flexible enough to account for high or very low over-dispersion).

The GAM that we will use for this problem is as follows:

$$
Y_i \sim \text{NegBinom}(D_i \mu_i, \tau_i)\\
\text{log}(\mu_i) = \beta_0 + \beta_1 \cdot \text{cos}\left(\frac{2\pi t_i}{365.25} \right) + \beta_2 \cdot \text{sin}\left(\frac{2\pi t_i}{365.25} \right) + \beta_3 \cdot \text{cos}\left(\frac{2\pi t_i}{182.625} \right) + \beta_4 \cdot \text{sin}\left(\frac{2\pi t_i}{182.625} \right) + f(t_i, \alpha)
$$

where:

-   $Y_i$ is the number of motorcycle deaths in month $i$

-   $\mu_i$ is the expected number of motorcycle deaths in month $i$

-   $D_i$ is the number of days in month $i$. We use `logMonthDays` as an offset in our model to account for differences in exposure time.

-   The $\beta_i$'s are the coefficients of the seasonal effects for 6 and 12 month intervals.

-   $f$ is the smooth effect for date (smooth function over time).

-   $\alpha$ is the smoothing coefficient.

# Part 2				

Show R code which fits this model using the mgcv package. (2 points)

```{r, echo=TRUE, message=FALSE, warning=FALSE}
motorcycleGAM <- mgcv::gam(
  update.formula(killed ~ offset(logMonthDays) +  s(dateInt, k=50),
  Pmisc::seasonalFormula(period=365.25, harmonics=1:2, var='dateInt')),
  data=x,
  family=nb(link='log'),
  method='ML'
)

# motorcycleGAM <- mgcv::gam(killed ~ month + offset(logMonthDays) + s(dateInt, k=50),
#                    data=x,
#                    family=nb,
#                    method='ML'
#                    )
```

# Part 3				

Produce a figure similar to Figure fig. 1 which is able to visualize the trend estimated from the
motorcycle data. You’re marked on the figure looking professional (with clear labels and a caption) as
well as conveying the important statistical information (prediction intervals as well as point predictions).				

(4 points)

```{r, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

# plot the actual data points
plot(x$date, 
     x$killed,
     col='darkgray', 
     log='y',
     xlab='', 
     ylab='Motorcycle deaths',
     cex=0.4
     )

# plot seasonal effect
matlines(x$date,
         exp(do.call(cbind, predict(motorcycleGAM, se.fit=TRUE)) %*% Pmisc::ciMat(0.95)),
         lty=c(1,2,2), col='#377eb8'

)

# plot the wiggly line (trend over time)
newX = x
newX$month = 'Mar'
# newX$month <- factor('Mar', levels = levels(x$month))  # Ensure factor levels are the same
newX$logMonthDays = log(31)

# use linear function for the straight line
matlines(newX$date,
         exp(do.call(cbind, predict(motorcycleGAM, newX,
                                    exclude = attributes(res$terms)$term.labels[2:5],
                                    se.fit=TRUE)) %*% Pmisc::ciMat(0.95)),
         lty = c(1, 2, 2),
         col='coral3',
         lwd = c(1.5, 1, 1),
         alpha=0.5
         )

# matlines(newX$date,
#          exp(do.call(cbind, predict(motorcycleGAM, newX, 
#                                     se.fit=TRUE)) %*% Pmisc::ciMat(0.95)),
#          lty = c(1, 2, 2),
#          col='coral3',
#          lwd = c(1.5, 1, 1),
#          alpha=0.5
#          )
```

```{r}
mgcv::exclude.terms()
```

				

			

		

	

# 
			

		

	
