---
title: "sta303 a2 q2"
author: "Talia Fabregas"
date: "2025-03-28"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

# Provided code

```{r, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# download the data
heatUrl = "http://pbrown.ca/teaching/appliedstats/data/sableIsland.rds"
dir.create("cache", showWarnings = FALSE)
heatFile = file.path("cache", basename(heatUrl))

if (!file.exists(heatFile)) 
  download.file(heatUrl, heatFile)

x = readRDS(heatFile)
names(x) = gsub("[.]+C[.]", "", names(x))
x$Date = as.Date(x$Date)
x$month = as.numeric(format(x$Date, "%m"))
x$summer = x$month %in% 5:10
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
x[100, ]
x$dateInt = as.integer(x$Date)
x$yearFac = factor(format(x$Date, "%Y"))

xSub = x[x$summer & !is.na(x$Max.Temp), ]

# View(xSub)
# View(x)

library("mgcv")

# Model 1 (res1)
res1 = gam(update.formula(
  Max.Temp ~ s(dateInt, pc = as.integer(as.Date("1990/7/1")), k = 100) + 
  s(yearFac, bs = "re"), 
  Pmisc::seasonalFormula(period = 365.25, harmonics = 1:2, var = "dateInt")
), data = xSub, method = "ML", optimizer = "efs")

# Model 2 (res2)
res2 = gam(update.formula(
  Max.Temp ~ s(dateInt, pc = as.integer(as.Date("1990/7/1")), k = 4) + 
  s(yearFac, bs = "re"), 
  Pmisc::seasonalFormula(period = 365.25, harmonics = 1:2, var = "dateInt")
), data = xSub, method = "ML", optimizer = "efs")

# Model 3 (res3)
res3 = gam(update.formula(
  Max.Temp ~ s(dateInt, pc = as.integer(as.Date("1990/7/1")), k = 100), 
  Pmisc::seasonalFormula(period = 365.25, harmonics = 1:2, var = "dateInt")
), data = xSub, method = "ML", optimizer = "efs")

# Output the formula of res1
res1$formula
res2$formula
res3$formula
```

# Part 1

Equations for `res1`

$$
Y_i \sim \mathcal{N}(\mu_i, \sigma^2) \\
\mu_i = \beta_0 + \beta_1 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_2 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_3 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + \beta_4 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + \\
 f_1(\text{dateInt}_i) + f_2(\text{yearFac}_i)
$$ where:

-   $Y_i$ is the Max Temp recorded in month $i$

-   $\beta_1$, $\beta_2$ are the coefficients of the cosine and sine
    terms of the 12-month cycle

-   $\beta_3$ and $\beta_4$ are the coefficients of the cosine and sine
    terms of the 6-month cycle

-   $f_1$ is the smooth trend over time (dateInt), modeled using a
    spline with k=100 knots

-   $f_2$ is the year random effect which accounts for variability
    between years. The use of `bs = "re"` in the code tells `mgcv::gam`
    to treat it as a random effect, and not a smoothing term.

Equations for `res2`$$
Y_i \sim \mathcal{N}(\mu_i, 0) \\
\mu_i = \beta_0 + \beta_1 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_2 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_3 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + \beta_4 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) \\
+ f_1(\text{dateInt}_i) + f_2(\text{yearFac}_i)
$$ where:

-   $Y_i$ is the Max Temp recorded in month $i$

-   $\beta_1$, $\beta_2$ are the coefficients of the cosine and sine
    terms of the 12-month cycle

-   $\beta_3$ and $\beta_4$ are the coefficients of the cosine and sine
    terms of the 6-month cycle

```         
-    $f_1$ is the smooth trend over time (dateInt), modeled using a
    spline with k=4 knots. The number of knots, k is the key
    difference between `res1` and `res2`
```

-   $f_2$ is the year random effect which accounts for variability
    between years. The use of `bs = "re"` in the code tells `mgcv::gam`
    to treat it as a random effect, and not a smoothing term.

Equations for `res3` $$
Y_i \sim \mathcal{N}(\mu_i, 0) \\
\mu_i = \beta_0 + \beta_1 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_2 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_3 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + \beta_4 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + f(\text{dateInt}_i) 
$$ where:

-   $Y_i$ is the Max Temp recorded in month $i$

-   $\beta_1$, $\beta_2$ are the coefficients of the cosine and sine
    terms of the 12-month cycle

-   $\beta_3$ and $\beta_4$ are the coefficients of the cosine and sine
    terms of the 6-month cycle

-   $f$ is the smooth trend over time (dateInt), modeled using a spline
    with k=100 knots.

Differences between the models in `res1`, `res2`, and `res3`:

# Part 2

# Part 3

# Part 4

Re-fitting the model using only data from 1995 and earlier

Comparable to the covid excess ontario deaths shown in lecture — what
would things have looked like if trends up to 1995 continued???

looking at excess warming, analogous to excess mortality???

```{r}
# filter data to only include before 1996
data_1995 <- subset(x, dateInt <= as.integer(as.Date("1995-12-31")))
```

```{r}
# refit the model
tempGAM1996 = gam(
  update.formula(
    Max.Temp ~ s(dateInt, pc = as.integer(as.Date("1990-07-01")), k = 100) + 
      s(yearFac, bs = "re"), 
  Pmisc::seasonalFormula(period = 365.25, harmonics = 1:2, var = "dateInt")), 
  data = data_1995
  )
```

```{r}
# Create a new dataset for forecasting

# dates to forecast based on pre-1996 data
forecast_dates <- seq(as.Date("1996-01-01"), as.Date("2025-12-31"), by = "day")

# create the forecast dataframe
forecast_data <- data.frame(
  date = forecast_dates,
  dateInt = as.integer(forecast_dates),
  yearFac = factor(format(forecast_dates, "%Y"))
)

# Predict
# dont think i need this line
# forecast_data$forecast <- predict(tempGAM1996, newdata = forecast_data)
```

```{r}
# forecast temperatures based on pre-1996 data
# what would have happened without 1996 global warming
forecast2025 = do.call(cbind,
                       predict(tempGAM1996, forecast_data, se.fit=TRUE)
                       ) %*% Pmisc::ciMat()

forecast_data$forecast <- forecast2025[, 1]  # Mean forecast
forecast_data$lower <- forecast2025[, 2]     # Lower 95% 
forecast_data$upper <- forecast2025[, 3]     # Upper 95% 
```
```{r, echo=FALSE, warning=FALSE, message=FALSE, include=TRUE}
# Convert dateInt to Date format
dates <- as.Date(comparison$dateInt, origin = "1970-01-01")

# Plot actual data first
plot(dates, comparison$Max.Temp, col = "gray", pch = 16, cex = 0.4,
     xlab = "Date", ylab = "Max Temperature (°C)",
     main = "Forecasted vs Actual Max Temperatures (1996-2025)")

# Add forecast trend line
lines(dates, comparison$forecast, col = "magenta4", lwd = 1.5)

# Add 95% prediction intervals
lines(dates, comparison$lower, col = "magenta4", lwd = 1, lty = 2)  # Lower bound
lines(dates, comparison$upper, col = "magenta4", lwd = 1, lty = 2)  # Upper bound

polygon(c(dates, rev(dates)), 
        c(comparison$lower, rev(comparison$upper)), 
        col = rgb(0.7, 0.7, 0.7, 0.3), border = NA)

# Add legend
legend("topright", legend = c("Actual", "Forecast", "95% Prediction Interval"),
       col = c("gray", "magenta4", "green"),
       pch = c(16, NA, NA), lty = c(NA, 1, 2), lwd = c(NA, 2, 1), bty = "n")
```

```{r}
library(ggplot2)

# Merge forecast with actual data
actual_data <- subset(x, dateInt >= as.integer(as.Date("1996-01-01"))) # get the actual data
comparison <- merge(forecast_data, actual_data, by = "dateInt", all.x = TRUE)

ggplot(comparison, aes(x = as.Date(dateInt, origin = "1970-01-01"))) +
  geom_line(aes(y = forecast, color = "Forecast"), size = 1) +          # Forecast trend line
  geom_point(aes(y = Max.Temp, color = "Actual"), size = 1.5, alpha = 0.5) +  # Actual data points
  labs(title = "Forecast Trend Line with Actual Data Points",
       x = "Date", y = "Temperature (°C)") +
  scale_color_manual(values = c("Forecast" = "purple4", "Actual" = "lightblue3")) +
  theme_minimal()
```

```{r}
# Prepare the data
comparison <- merge(forecast_data, actual_data, by = "dateInt", all.x = TRUE)

# Convert dateInt to Date format
dates <- as.Date(comparison$dateInt, origin = "1970-01-01")

# Plot the forecast trend line
matplot(dates, comparison$forecast, type = "l", col = "magenta4", lwd = 2, lty = 1,
        xlab = "Date", ylab = "Max Temperature (°C)",
        main = "Forecast Trend Line with Actual Data Points")

# Add actual data points
points(dates, comparison$Max.Temp, col = "lightblue3", pch = 16, cex = 0.7)

# Add legend
legend("topright", legend = c("Forecast", "Actual"), 
       col = c("magenta4", "lightblue3"), lty = 1, pch = c(NA, 16), bty = "n")
```
```{r}
# Prepare the data
comparison <- merge(forecast_data, actual_data, by = "dateInt", all.x = TRUE)

# Convert dateInt to Date format
dates <- as.Date(comparison$dateInt, origin = "1970-01-01")

# Plot the actual data first
plot(dates, comparison$Max.Temp, col = "lightblue3", pch = 16, cex = 0.7,
     xlab = "Date", ylab = "Max Temperature (°C)",
     main = "Forecast Trend Line on Top of Actual Data")

# Overlay the forecast trend line
lines(dates, comparison$forecast, col = "magenta4", lwd = 2)

# Add legend
legend("topright", legend = c("Actual", "Forecast"),
       col = c("lightblue3", "magenta4"), pch = c(16, NA), lty = c(NA, 1), lwd = 2, bty = "n")

```





```{r, echo=FALSE, warning=FALSE, message=FALSE}

# fit model up to December 31 1995
# next step is to forecast
# tempGAM1996 = gam(res1$formula,
#                   method='ML',
#                   data= x[x$Date <= as.Date("1995-12-31") & !is.na(x$Max.Temp), ],
#                   optimizer='efs'
#                   ) 

x4 = x[x$Date <= as.Date("1995-12-31") & !is.na(x$Max.Temp), ]

tempGAM1996 = gam(update.formula(
      Max.Temp ~ s(dateInt, pc = as.integer(as.Date("1990/7/1")), k = 100) + 
      s(yearFac, bs = "re"), 
      Pmisc::seasonalFormula(period = 365.25, harmonics = 1:2, var = "dateInt")
    ),
    data=x4,
    method='ML',
    optimizer='efs'
)
```

```{r}
# forecast
jan_1_1996 = as.Date('1996/1/1')
jan_1_2025 = as.Date('2025/1/1')
X_forecast = data.frame(
  date = seq(jan_1_1996, jan_1_2025, by='month'),
  month='Mar')
X_forecast$dateInt = as.integer(X_forecast$date)

pred4 = do.call(
  cbind,
  predict(tempGAM1996, X_forecast, se.fit=TRUE)
  ) %*% Pmisc::ciMat()
```

```{r}
# compare model forecast to observed data
matplot(X_forecast$date, pred4,
        lty=c(1,2,2),
        col='red',
        xlim = c(jan_1_1996, jan_1_2025),
        xlab='',
        ylab = 'temperature'
        )

# plot the actual data points
points(xSub$Date, xSub$Max.Temp, cex=0.3)

# plot the trend?????
abline(v=max(tempGam1996$model$dateInt)) 
```
