---
title: "sta303 a2 q2"
author: "Talia Fabregas"
date: "2025-03-28"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

# Provided code

```{r, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# download the data
heatUrl = "http://pbrown.ca/teaching/appliedstats/data/sableIsland.rds"
dir.create("cache", showWarnings = FALSE)
heatFile = file.path("cache", basename(heatUrl))

if (!file.exists(heatFile)) 
  download.file(heatUrl, heatFile)

x = readRDS(heatFile)
names(x) = gsub("[.]+C[.]", "", names(x))
x$Date = as.Date(x$Date)
x$month = as.numeric(format(x$Date, "%m"))
x$summer = x$month %in% 5:10
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
x[100, ]
x$dateInt = as.integer(x$Date)
x$yearFac = factor(format(x$Date, "%Y"))

xSub = x[x$summer & !is.na(x$Max.Temp), ]

# View(xSub)
# View(x)

library("mgcv")

# Model 1 (res1)
res1 = gam(update.formula(
  Max.Temp ~ s(dateInt, pc = as.integer(as.Date("1990/7/1")), k = 100) + 
  s(yearFac, bs = "re"), 
  Pmisc::seasonalFormula(period = 365.25, harmonics = 1:2, var = "dateInt")
), data = xSub, method = "ML", optimizer = "efs")

# Model 2 (res2)
res2 = gam(update.formula(
  Max.Temp ~ s(dateInt, pc = as.integer(as.Date("1990/7/1")), k = 4) + 
  s(yearFac, bs = "re"), 
  Pmisc::seasonalFormula(period = 365.25, harmonics = 1:2, var = "dateInt")
), data = xSub, method = "ML", optimizer = "efs")

# Model 3 (res3)
res3 = gam(update.formula(
  Max.Temp ~ s(dateInt, pc = as.integer(as.Date("1990/7/1")), k = 100), 
  Pmisc::seasonalFormula(period = 365.25, harmonics = 1:2, var = "dateInt")
), data = xSub, method = "ML", optimizer = "efs")

# Output the formula of res1
res1$formula
res2$formula
res3$formula
```

# Part 1

Equations for `res1`

$$
Y_i \sim \mathcal{N}(\mu_i, \sigma^2) \\
\mu_i = \beta_0 + \beta_1 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_2 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_3 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + \beta_4 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + f_1(\text{dateInt}_i) + f_2(\text{yearFac}_i)
$$ where:

-    $Y_i$ is the Max Temp recorded in month $i$

-    $\beta_1$, $\beta_2$ are the coefficients of the cosine and sine
    terms of the 12-month cycle

-    $\beta_3$ and $\beta_4$ are the coefficients of the cosine and sine
    terms of the 6-month cycle

-    $f_1$ is the smooth trend over time (dateInt), modeled using a
    spline with k=100 knots

-    $f_2$ is the year random effect which accounts for variability
    between years. The use of `bs = "re"` in the code tells `mgcv::gam`
    to treat it as a random effect, and not a smoothing term.

Equations for `res2`$$
Y_i \sim \mathcal{N}(\mu_i, 0) \\
\mu_i = \beta_0 + \beta_1 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_2 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_3 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + \beta_4 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + f_1(\text{dateInt}_i) + f_2(\text{yearFac}_i)
$$ where:

-    $Y_i$ is the Max Temp recorded in month $i$

-   $\beta_1$, $\beta_2$ are the coefficients of the cosine and sine
    terms of the 12-month cycle

-    $\beta_3$ and $\beta_4$ are the coefficients of the cosine and sine
    terms of the 6-month cycle

    -    $f_1$ is the smooth trend over time (dateInt), modeled using a
        spline with k=4 knots. The number of knots, k is the key
        difference between `res1` and `res2`

-    $f_2$ is the year random effect which accounts for variability
    between years. The use of `bs = "re"` in the code tells `mgcv::gam`
    to treat it as a random effect, and not a smoothing term.

Equations for `res3` $$
Y_i \sim \mathcal{N}(\mu_i, 0) \\
\mu_i = \beta_0 + \beta_1 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_2 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{365.25} \right) + \beta_3 \cdot \text{cos}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + \beta_4 \cdot \text{sin}\left(\frac{2\pi \text{dateInt}_i}{182.625} \right) + f(\text{dateInt}_i) 
$$ where:

-    $Y_i$ is the Max Temp recorded in month $i$

-   $\beta_1$, $\beta_2$ are the coefficients of the cosine and sine
    terms of the 12-month cycle

-    $\beta_3$ and $\beta_4$ are the coefficients of the cosine and sine
    terms of the 6-month cycle

-    $f$ is the smooth trend over time (dateInt), modeled using a spline
    with k=100 knots.

# Part 2

# Part 3

# Part 4

Re-fitting the model using only data from 1995 and earlier

Comparable to the covid excess ontario deaths shown in lecture — what
would things have looked like if trends up to 1995 continued???

looking at excess warming, analogous to excess mortality

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# fit model up to December 31 1995
tempGAM1996 = gam(res1$formula,
                  method='ML',
                  data= x[x$Date <= as.Date("1995-12-31") & !is.na(x$Max.Temp), ]
                  ) 
```

```{r}
jan_1_1996 = as.Date('1996/1/1')
jan_1_2025 = as.Date('2025/1/1')
X_forecast = data.frame(
  date = seq(jan_1_1996, jan_1_2025, by='month'),
  month='Mar')
X_forecast$dateInt = as.integer(X_forecast$date)
```
