---
title: "STA303 Assignment 2"
author: 
- Talia Fabregas
- Kasandra Tworzynaski
date: \today
date-format: long
number-sections: true
output: pdf_document
---

# Q1: Motorcycle deaths

```{r pmisc}
if(!requireNamespace("Pmisc", quietly=TRUE)) {
	install.packages("Pmisc", repos='http://r-forge.r-project.org')
}
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
library(Hmisc)
library(Pmisc)
# install.packages("mcgv")
library(mgcv)
library(kableExtra)
library(janitor)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# retrieving the data
theUrl = "http://pbrown.ca/teaching/appliedstats/data/motorcycle.rds"
theFile = basename(theUrl)
if (!file.exists(theFile)) download.file(theUrl, theFile)
x = readRDS(theFile)

# preparing the data
x$dateInt = as.integer(x$date)
x$logMonthDays = log(Hmisc::monthDays(x$date))
x$month = factor(format(x$date, "%b"), levels = format(ISOdate(2000,
+   1:12, 1), "%b"))
res = glm(killed ~ offset(logMonthDays) + dateInt + month, data = x, 
          family = poisson(link = "log")
          )
newdata = data.frame(date = seq(as.Date("1975/1/1"), as.Date("2030/1/1"), by = "month"))
newdata$dateInt = as.integer(newdata$date)
newdata$logMonthDays = log(30)
newdata$month = "Mar"
pred1 = predict(res, newdata)
newdata$month = format(newdata$date, "%b")
pred2 = predict(res, newdata)
plot(x$date, x$killed, cex = 0.2, log = "y", xlab = "", ylab = "")
matlines(newdata$date, exp(cbind(pred1, pred2)), lty = 1)
```

```{r}
x
```

1.  Write down, in equations not R code, a generalized additive model suitable for this problem. Explain each of the parts of the model and give a rationale for them (i.e. “The response variable is Gamma distributed because the number of deaths must be positive”). (4 points)

    Do we have to check for overdispersion or is this already telling us it is poission?

    Number of motorcylcle deaths every month from 1980 to 2025

    An appropriate generalized additive model

    We chose a poisson because the response variable, the number of motorcycle deaths in month i, is a count variable.

    Covariate Xi is month

    Beta is the fixed effect of month
  $$
    Y_i  \sim Pois(D_i \lambda_i) \\
    \text{log}(\lambda_i) = X_i \beta + U(t_i) \\
    U(.) \sim IWP_2(\sigma)
  $$
The response variable, $Y_i$, is Poisson because the number of motorcycle deaths is a count variable. $D_i$ is the number of deaths in month $i$ and $\lambda_i$ is the intensity. 

2. Show R code to fit the model using the `mcgv` package
```{r, echo=TRUE, message=FALSE, warning=FALSE}
model1 <- gam(killed ~ s(dateInt) + month + offset(logMonthDays),
              data=x,
              family=poisson,
              method = 'ML')
```

3. Produce a figure similar to Figure fig. 1 which is able to visualize the trend estimated from the motorcycle data. You’re marked on the figure looking professional (with clear labels and a caption) as well as conveying the important statistical information (prediction intervals as well as point predictions).
(4 points)

```{r, fig.cap="There has been an overall decline in deaths overtime, since the highest levels were seen around 1980. The red line trend shows there are seasonal variations in the number of deaths in the month. The points in black appear to be modeled well by the trend in red"}
#3 
# First prediction: using March as the month
newdata$month <- "Mar"
predx <- predict(model1, newdata)

# Second prediction: using the actual month of each date
newdata$month <- format(newdata$date, "%b")
predy <- predict(model1, newdata)

# Plot original data
plot(
  x$date, x$killed, 
  cex = 0.2, log = "y", 
  xlab = "Year", ylab = "Number of Deaths"
)

# Add predicted lines to the plot
matlines(
  newdata$date, exp(cbind(predx, predy)), 
  lty = 1
)
```

# Q2: Heat

```{r, echo=FALSE, message=FALSE, warning=FALSE}
x[100, ]
```

```{r. echo=FALSE, message=FALSE, warning=FALSE}
x$dateInt = as.integer(x$Date)
x$yearFac = factor(format(x$Date, "%Y"))
# xSub = x[x$summer & !is.na(x$Max.Temp), ]
# res1 = gam(update.formula(Max.Temp ~ s(dateInt, pc = as.integer(as.Date("1990/7/1")),
# +   k = 100) + s(yearFac, bs = "re"), Pmisc::seasonalFormula(period = 365.25,
# +   harmonics = 1:2, var = "dateInt")), data = xSub, method = "ML",
# +   optimizer = "efs")
```
