---
title: "STA303 Assignment 2"
author: 
- Talia Fabregas
- Kasandra Tworzynaski
date: \today
date-format: long
number-sections: true
output: pdf_document
---

# Motorcycle deaths

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
library(Hmisc)
library(Pmisc)
# install.packages("mcgv")
library(mgcv)
library(kableExtra)
library(janitor)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# retrieving the data
theUrl = "http://pbrown.ca/teaching/appliedstats/data/motorcycle.rds"
theFile = basename(theUrl)
if (!file.exists(theFile)) download.file(theUrl, theFile)
x = readRDS(theFile)

# preparing the data
x$dateInt = as.integer(x$date)
x$logMonthDays = log(Hmisc::monthDays(x$date))
x$month = factor(format(x$date, "%b"), levels = format(ISOdate(2000,
+   1:12, 1), "%b"))
res = glm(killed ~ offset(logMonthDays) + dateInt + month, data = x, family = poisson(link = "log"))
newdata = data.frame(date = seq(as.Date("1975/1/1"), as.Date("2030/1/1"), by = "month"))
newdata$dateInt = as.integer(newdata$date)
newdata$logMonthDays = log(30)
newdata$month = "Mar"
pred1 = predict(res, newdata)
newdata$month = format(newdata$date, "%b")
pred2 = predict(res, newdata)
plot(x$date, x$killed, cex = 0.2, log = "y", xlab = "", ylab = "")
matlines(newdata$date, exp(cbind(pred1, pred2)), lty = 1)
```

```{r}
x
```

1.  Write down, in equations not R code, a generalized additive model suitable for this problem. Explain each of the parts of the model and give a rationale for them (i.e. “The response variable is Gamma distributed because the number of deaths must be positive”). (4 points)

    Do we have to check for overdispersion or is this already telling us it is poission?

    Number of motorcylcle deaths every month from 1980 to 2025

    An appropriate generalized additive model

    We chose a poisson because the response variable, the number of motorcycle deaths in month i, is a count variable.

    Covariate Xi is month

    Beta is the fixed effect of month
  $$
    Y_i  \sim Pois(D_i \lambda_i) \\
    G(\lambda_i) = X_i \beta + U(t_i) \\
    U(.) \sim IWP_2(\sigma)
  $$

2. Show R code to fit the model using the `mcgv` package
3. 


# Heat

```{r, echo=FALSE, message=FALSE, warning=FALSE}
x[100, ]
```

```{r. echo=FALSE, message=FALSE, warning=FALSE}
x$dateInt = as.integer(x$Date)
x$yearFac = factor(format(x$Date, "%Y"))
# xSub = x[x$summer & !is.na(x$Max.Temp), ]
# res1 = gam(update.formula(Max.Temp ~ s(dateInt, pc = as.integer(as.Date("1990/7/1")),
# +   k = 100) + s(yearFac, bs = "re"), Pmisc::seasonalFormula(period = 365.25,
# +   harmonics = 1:2, var = "dateInt")), data = xSub, method = "ML",
# +   optimizer = "efs")
```
